---
title: "More Information"
author: "Chichi Wu and Megan Varnum"
date: "Last updated on `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: kable
---

<!--
Comments in HTML are like this! 
-->

# More Information

## Average Age Overtime


```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Set R chunk default options:
# -do not output all R messages in html report
# -do not output all R warnings in html report
# -set default figure widths and heights
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE, fig.width = 16/2, fig.height = 9/2
)
```


```{r, echo = FALSE}
#load packages
library(tidyverse)

#load datasets

athlete_events <- read_csv("athlete_events.csv")  %>% 
  mutate(Sex = ifelse(Sex == "F", "Women", "Men"))
noc_regions <- read_csv("noc_regions.csv")

#data wrangling
athlete_all <- athlete_events %>% 
  filter(Sport == "Swimming" | Sport == "Gymnastics" | Sport == "Athletics") %>% 
  group_by(Year, Sex, Sport) %>% 
  summarize(
    avg_height = mean(Height, na.rm = TRUE),
    avg_weight = mean(Weight, na.rm = TRUE),
    avg_age = mean(Age, na.rm = TRUE), 
    avg_BMI = mean(Weight, na.rm = TRUE) / ((mean(Height, na.rm = TRUE) / 100)*(mean(Height, na.rm = TRUE) / 100)))%>% 
  filter(avg_height != "NaN" & avg_weight != "NaN" & avg_age != "NaN")

#average age visualization

ggplot(data = athlete_all, 
       mapping = aes(x = Year, 
                     y = avg_age, 
                     color = Sex)) +
  #geom_point() +
  geom_smooth(se = FALSE, span = 0.75) +
  facet_wrap(~Sport) +
  labs(y = "Mean Athlete Age", 
       title = "Olympian Age Over the Years:\nSwimming, Gymnastics, Track & Field") +
  theme(legend.title = element_blank()) +
  xlim(1900, 2016)

```
 
 
In looking at the average age of Olympians across time, an overall trend shows that male athletes are consistently older than female athletes across all three sports. The reason for why females peak earlier in age than males may be attributed to societal pressures that women face outside of sports, such as dating and starting a family. Additionally, for gymnastics, the priority for female athletes is flexibility, whereas for male athletes it is strength. Thus, a younger age demographic corresponds to higher flexibility in women. It can be noted that in recent years, from around 1980, the average age has been increasing across all sports. This perhaps is due to a shift in the dialogue around women and sports and a new preference for women to also exhibit exceptional strength. However, the age range is still very low, the average for gymnastics being around twenty in 2012. This brings up many ethical questions, especially in light of the #MeToo movement and the scandal within the US team. When young girls are expected to compete professionally, what kinds of protection are enacted?


## Swimming and Track BMI by Event

```{r, echo = FALSE}
#comparing runners and swimmers BMI accross events

#400 or less
track_sprint_events <- list(
  "Athletics Men's 100 metres", 
  "Athletics Women's 100 metres", 
  "Athletics Men's 200 metres", 
  "Athletics Women's 200 metres", 
  "Athletics Men's 400 metres", 
  "Athletics Women's 400 metres", 
  "Athletics Men's 4 x 100 metres Relay",
  "Athletics Women's 4 x 100 metres Relay",
  "Athletics Men's 4 x 400 metres Relay",
  "Athletics Women's 4 x 400 metres Relay"
  )

#steeple excluded
track_hurdle_events <- list(
  "Athletics Men's 110 metres Hurdles", 
  "Athletics Women's 100 metres Hurdles",
  "Athletics Men's 400 metres Hurdles", 
  "Athletics Women's 400 metres Hurdles"
)

#800 and 1500
track_middle_events <- list(
  "Athletics Men's 1,500 metres",
  "Athletics Women's 1,500 metres",
  "Athletics Men's 800 metres",
  "Athletics Women's 800 metres"
)

#3k and above, steeple included
track_distance_events <- list(
  "Athletics Women's 3,000 metres",
  "Athletics Men's 5,000 metres",
  "Athletics Women's 5,000 metres",
  "Athletics Men's 3,000 metres Steeplechase",
  "Athletics Women's 3,000 metres Steeplechase",
  "Athletics Men's 10,000 metres",
  "Athletics Women's 10,000 metres"
  )

track_marathon_events <- list(
  "Athletics Men's Marathon",
  "Athletics Women's Marathon"
)

track_walk_events <- list(
  "Athletics Women's 10 kilometres Walk",
  "Athletics Men's 20 kilometres Walk",
  "Athletics Women's 20 kilometres Walk",
  "Athletics Men's 50 kilometres Walk",
  "Athletics Women's 50 kilometres Walk"
)


swim_free_sprint_events <- list(
  "Swimming Women's 50 metres Freestyle",
  "Swimming Women's 400 metres Freestyle",
  "Swimming Women's 4 x 200 metres Freestyle Relay",
  "Swimming Women's 4 x 100 metres Freestyle Relay",
  "Swimming Women's 200 metres Freestyle",
  "Swimming Women's 100 metres Freestyle",
  "Swimming Men's 50 metres Freestyle",
  "Swimming Men's 400 metres Freestyle",
  "Swimming Men's 4 x 200 metres Freestyle Relay",
  "Swimming Men's 4 x 100 metres Freestyle Relay",
  "Swimming Men's 200 metres Freestyle",
  "Swimming Men's 100 metres Freestyle"
)

swim_free_distance_events <- list(
  "Swimming Men's 1,500 metres Freestyle",
  "Swimming Women's 800 metres Freestyle",
  "Swimming Women's 1,500 metres Freestyle",
  "Swimming Men's 800 metres Freestyle",
  "Swimming Men's 10 kilometres Open Water",
  "Swimming Women's 10 kilometres Open Water"
)

swim_butterfly_events <- list(
  "Swimming Women's 200 metres Butterfly",
  "Swimming Women's 100 metres Butterfly",
  "Swimming Men's 200 metres Butterfly",
  "Swimming Men's 100 metres Butterfly"
)

swim_breast_events <- list(
  "Swimming Women's 200 metres Breaststroke",
  "Swimming Women's 100 metres Breaststroke",
  "Swimming Men's 200 metres Breaststroke",
  "Swimming Men's 100 metres Breaststroke"
)

swim_back_events <- list(
  "Swimming Women's 200 metres Backstroke",
  "Swimming Women's 100 metres Backstroke",
  "Swimming Men's 200 metres Backstroke",
  "Swimming Men's 100 metres Backstroke"
)

swim_IM_events <- list(
  "Swimming Women's 400 metres Individual Medley",
  "Swimming Women's 4 x 100 metres Medley Relay",
  "Swimming Women's 200 metres Individual Medley",
  "Swimming Men's 400 metres Individual Medley",
  "Swimming Men's 4 x 100 metres Medley Relay",
  "Swimming Men's 200 metres Individual Medley"
)

```

```{r, echo = FALSE}
track_event_order <- c("Sprint",
                 "Hurdles",
                 "Walk",
                 "Middle",
                 "Distance",
                 "Marathon")

swim_event_order <- c("IM",
                      "Breast",
                      "Sprint",
                      "Butterfly",
                      "Back",
                      "Distance")

```

```{r, echo = FALSE}
runner_BMI <- athlete_events %>% 
  filter(Sport == "Athletics", Year > 1990) %>% 
  mutate(BMI = Weight/((Height/100)*(Height/100))) %>% 
  #filter(grepl("0", Event)) %>% 
  mutate(event_catagory = 
           ifelse(Event %in% track_sprint_events, "Sprint", 
                  ifelse(Event %in% track_distance_events, "Distance", 
                         ifelse(Event %in% track_middle_events, "Middle",
                                ifelse(Event %in% track_hurdle_events, "Hurdles", 
                                       ifelse(Event %in% track_walk_events, "Walk", 
                                              ifelse(Event %in% track_marathon_events, "Marathon",
                                                     "Other"))))))) %>% 
  filter(event_catagory != "Other") %>% #get rid of field events
  mutate(event_catagory = factor(event_catagory, levels = track_event_order))

swimmer_BMI <- athlete_events %>% 
  filter(Sport == "Swimming", Year > 1990) %>% 
  mutate(BMI = Weight/((Height/100)*(Height/100))) %>% 
  mutate(event_catagory = 
           ifelse(Event %in% swim_free_distance_events, "Distance",
                  ifelse(Event %in% swim_free_sprint_events, "Sprint",
                         ifelse(Event %in% swim_butterfly_events, "Butterfly",
                                ifelse(Event %in% swim_breast_events,"Breast",
                                       ifelse(Event %in% swim_back_events, "Back",
                                              ifelse(Event %in% swim_IM_events, "IM", "Other"))))))) %>% 
  mutate(event_catagory = factor(event_catagory, levels = swim_event_order))
```

```{r, echo = FALSE}
ggplot(data = runner_BMI, mapping = aes(x = factor(event_catagory), y = BMI, fill = event_catagory)) +
  facet_wrap(~Sex) +
  geom_boxplot() +
  labs(x = "Type of Event",
       y = "Avergae Athlete BMI",
       title = "Olympian's BMI by Event: Runners")+
  ylim(14, 35) +
  scale_fill_brewer(palette = "Set3") +
  theme(legend.position = "none") 

ggplot(data = swimmer_BMI, mapping = aes(x = factor(event_catagory), y = BMI, fill = event_catagory)) +
  facet_wrap(~Sex) +
  geom_boxplot() +
  labs(x = "Type of Event",
       y = "Avergae Athlete BMI",
       title = "Olympian's BMI by Event: Swimmers") + 
  ylim(14, 35)  +
  scale_fill_brewer(palette = "Set3") +
  theme(legend.position = "none")
```


Comparing the BMIs of runners and swimmers across event categories (distance, sprints, etc), we found significantly more consistency in swimmers. There appears to be a single ideal BMI for all swimmers, whereas a runner’s BMI is fine-tuned depending on their event. In runners, shorter events favor higher BMIs. Land athletes face a tradeoff between strength and endurance -- although large muscles equate to more power, they require additional oxygen transport. Swimmers do not face the same dilemma because drag is the primary limitation to performance instead of weight.^[ Hallman, Kevin. “Elite Swimmers Share Same Ideal BMI Across Different Race Distances.” SwimSwam, July 14th, 2018, https://swimswam.com/elite-swimmers-share-same-ideal-bmi-across-different-race-distances/] 


## Age of Medalists in Gymnastics 
```{r, echo = FALSE}
#wrangling
athlete_gymnastics <- athlete_events %>% 
  filter(Sport == "Gymnastics")

gymnastics_medal_winners <- athlete_gymnastics %>% 
    filter(Medal == "Silver" | Medal == "Bronze" | Medal == "Gold") %>% 
    filter(Year > 1999)
```

```{r, echo = FALSE}
ggplot(data = gymnastics_medal_winners, 
       mapping = aes(x = Age, fill = Sex)) + 
  geom_histogram(bins = 8,
                 aes(y = (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..])) + 
  facet_wrap(~Sex) +
  labs(x = "Age of Medal Winner", 
       y = "Proportion", 
       title = "Age of Medal Winners for Olympic Gymnastics: 2000-2016") +
  scale_fill_manual(values = c("#fb8072", "#80b1d3")) +
  theme(legend.position = "none")
```

Given the age of the medalists in gymnastics, it is shown that the histograms are right skewed for both men and women. The skew for female gymnasts, however, is a lot more considerable. Thus, this shows that younger gymnasts tend to perform better overall, and younger female gymnasts significantly perform better and are preferred. 

